Self-Supervised Learning of Generalized Policies in adversarial domains

### 2021-07-23 doc/paper/paper.tex
doc update

Generalized planning is concerned with the computation of general policies that solve multiple instances of a planning domain all at once. It has been recently shown that these policies can be computed in two steps: first, a suitable abstraction in the form of a qualitative numerical planning problem (QNP) is learned from sample plans, then the general policies are obtained from the learned QNP using a planner. In this work, we introduce an alternative approach for computing more expressive general policies which does not require sample plans or a QNP planner. The new formulation is very simple and can be cast in terms that are more standard in machine learning: a large but finite pool of features is defined from the predicates in the planning examples using a general grammar, and a small subset of features is sought for separating “good” from “bad” state transitions, and goals from non-goals. The problems of finding such a “separating surface” while labeling the transitions as “good” or “bad” are
jointly addressed as a single combinatorial optimization problem expressed as a Weighted Max-SAT problem. The advantage of looking for the simplest policy in the given feature space that solves the given examples, possibly non-optimally, is that many domains have no general, compact policies that are optimal. The approach yields general policies for a number of benchmark adversarial domains, even when non-determinism exists in the adverary turn. We use a uniform grammar for all the grid domains capable of capturing the domain representations. These policies can be interpreted qualitatively, explained, and it can be shown mathematically that they will solve all possible instances within a class of problems. This is an advantage when making manual policy changes. We also show an example of how the explainable policies produced by our algorithms can be applied to real robots.

@@

The main advantage of the formulation of generalized planning based on QNPs is that it applies to standard relational domains where the pool of (ground) actions change from instance to instance. On the other hand, while the planning domain is assumed to be given, the QNP abstraction is not, and hence it has to be written by hand or learned. This is the approach of Bonet, Franc`es, and Geffner (2019) where generalized plans are obtained by learning the QNP abstraction from the domain representation and sample plans, and then solving the abstraction with a QNP planner.

In this work, we build on this thread but introduce an alternative approach for computing general policies that is simpler, yet more powerful. The learning problem is cast as a self-supervised classification problem where (1) a pool of features is automatically generated from a uniform grammar applied to the domain predicates, and (2) a small subset of features is sought for separating “good” from “bad” state transitions, and goals from non-goals. The problems of finding the “separating surface” while labeling the transitions as “good” or “bad” are addressed jointly as a single combinatorial optimization task solved with a Weighted Max-SAT solver. The approach yields general policies for a number of benchmark domains.

One of the advantages of having explainable policies is that it gives the engineer or researcher the possibility to modify them easily if necessary. For instance, if there are known exceptions within the domain. Although this has not been necessary in the domains discussed in this paper.

@@

A key question in generalized planning is how to represent general plans or policies when the different instances to be solved have different sets of objects and ground actions. One solution is to work with general features (functions) that have well defined values over any state of any possible domain instance, and think of general policies π as mappings from feature valuations into abstract actions that denote changes in the feature values (Bonet and Geffner 2018). In this work, we build on this intuition but avoid the introduction of abstract actions (Bonet and Geffner 2021).

@@

\subsection{Space Invaders}
\subsubsection{Description}
This domain is a simplified version of the original Space Invaders in which the Martians cannot shoot and cannot move below the penultimate row. In this domain the agent can $shoot$, move to the $right$ or to the $left$. The agent during his turn can do two actions, while the Martians can do one. Note that if the Martians and the agent have only one action per turn the agent cannot win from all possible states in the game, this is because of the situation where the agent stands in the Martian's column to shoot him on his next turn, but the next turn (the Martian's turn) the Martian can move away, this leads to an infinite loop if both are playing perfectly. That is, the agent must be able to move to the agent's column and then shoot. To win the agent has to kill all the martians, for this he must shoot each of them. The agent kills an opponent if he shoots while being in the same column, and he can only kill one for each shoot. The martians has three possible actions $left$, $right$, and $down$. All the adversaries move in the same direction. The agent cannot be destroyed in any way by the Martians, they can go down to the row above the agent. The agent cannot collide with the Martians as they cannot descend to the row above it. Note that the policy found is optimal and will solve all possible instances of the game. The policy is also optimal for a version of the game where the Martians destroy the agent when they reach the row above him.

@@

\subsection{Checkmate in n with Queen}
\subsubsection{Description}

\subsubsection{Training instances}
\begin{Verbatim}[fontsize=\footnotesize]
  # .  k  .  .  K #
  # .  .  .  .  . #
  # .  .  .  .  . #
  # Q  .  .  .  . #
  # .  .  .  .  . #
\end{Verbatim}

\subsubsection{Features}
\begin{Verbatim}[fontsize=\footnotesize]
  (CM) Atom[checkmate] [k=1]
  (SM) Atom[stalemate] [k=1]
  (QbK) Atom[white_queen_between_kings] [k=1]
  (KE) Atom[black_king_closed_in_edge_by_white_queen] [k=1]
  (KD) Dist[cell-has-white_king;adjacent_cell;cell-has-black_king] [k=4]
  (BKhA) Bool[And(cell-attaked-by-black_king,Not(cell-attaked-by-white))] [k=4]
  (A) Num[And(cell-attaked-by-black_king,Not(cell-attaked-by-white_queen))] [k=4]
  (WQadjBK_row) Bool[And(row-has-white_queen,Exists(adjacent_row,row-has-black_king)))] [k=6]
  (WQadjBK_col) Bool[And(col-has-white_queen,Exists(adjacent_col,col-has-black_king)))] [k=6]
  (WQadjBK) Bool[And(cell-has-white_queen,Exists(adjacent_cell,cell-has-black_king)))] [k=6]
  (WQadjWK) Bool[And(cell-has-white_queen,Exists(adjacent_cell,cell-has-white_king)))] [k=6]
\end{Verbatim}

\subsubsection{All Transition Features}
This features reason about how the value of the features can change in all possible successors.
\begin{Verbatim}[fontsize=\footnotesize]
  (CA) Bool[can(A, DEC)] [k=2]
  (CCM) bool[can(CM, INC)] [k=2]
\end{Verbatim}

\subsubsection{Feature Description}
$CM$. Checkmate\\
$SM$. Stalemate\\
$KE$. Black king is closed in an edge of the table by the white queen\\
$QbK$. White queen is somewhere between kings, rows or columns\\
$A$. The number of cells attaked by black king and not attacked by white queen\\
$BKhA$. Exist a cell attaked by Black king and no attacked by white pieces\\
$Qaw$. White queen is attaked and without protection, i.e. $WQadjBK$ AND $WQadjWK$\\
$WQadjBK_line$. White queen is in an adjacent line to the black king, i.e. $WQadjBK_row$ OR $WQadjWK_col$\\
$CA$. It is possible to decrease the number of cells attaked by black king and not attacked by white queen\\
$CCM$. It is possible to do checkmate in the current state

\subsubsection{Policy}
  \begin{Verbatim}[fontsize=\footnotesize]
   1. KE=0, CA>0 -> {A ↓}
   2. KE=0, CA=0 -> {A ·, KD ·}
   3. BKhA=0, WQadjBK_line>0 -> {KD ·, WQadjBK_line ·}
   4. CCM=0, KE>0 -> {KD ↓}
   5. CCM=1 -> {CM ↑}
  \end{Verbatim}

\subsubsection{Policy Implicits}
The next features has to be in that certain value after the transition for classifing the transition as good, taking into account the policy rules.
  \begin{Verbatim}[fontsize=\footnotesize]
   Qaw=0, Qbk>0, SM=0
  \end{Verbatim}

\subsubsection{Policy description}
Rules 1 and 2 cover the cases in which the king is not enclosed in an edge of the file. The first rule states that it is good to take transitions that decrease the number of squares available to the black king by using the queen. The second rule says that if the first rule is not applicable it is good to take transitions that do not alter the number of squares available to the black king, without moving the white king. The third rule refers to those cases where the black king had moved to its only free square, and classifies as good those transitions in which the queen is moved so that it is not stalemate and remains on the line adjacent to the black king. Rule four covers cases in which the black king is locked on an edge by the queen, and it is good to move the white king closer. finally, rule 5 says that if it is possible to checkmate it is good to do so. It should be noted that thanks to the implicit parts of the policy the king will never be left tied up and unprotected, stalemate will never occur and the white king will always be placed and remain in a line that is between the lines of the kings, whether rows or columns.

\subsubsection{Example of Game}
\begin{Verbatim}[fontsize=\footnotesize]
  K.... # K.... # K.... # ..... # ..... # ..... # ..... # ..... # ..... #
  ..Q.. # ..... # ..... # ..K.. # ..K.. # ..K.. # ..... # ..... # ..... #
  ..... # ..... # ..... # ..... # ..... # ..... # ..K.. # ..K.. # ..K.. #
  ..... # Q.... # Q.... # Q.... # Q.... # Q.... # Q.... # Q.... # ...Q. #
  ....k # ....k # ...k. # ...k. # ..k.. # ...k. # ...k. # ....k # ....k #
\end{Verbatim}

\subsection{Checkmate in n with Rook}
\subsubsection{Description}

\subsubsection{Training instances}
\begin{Verbatim}[fontsize=\footnotesize]
  # .  k  .  .  K #
  # .  .  .  .  . #
  # .  .  .  .  . #
  # R  .  .  .  . #
  # .  .  .  .  . #
\end{Verbatim}

\subsubsection{Features}
\begin{Verbatim}[fontsize=\footnotesize]
  (CM) Atom[checkmate] [k=1]
  (SM) Atom[stalemate] [k=1]
  (QbK) Atom[white_rook_between_kings] [k=1]
  (KE) Atom[black_king_closed_in_edge_by_white_rook] [k=1]
  (KD) Dist[cell-has-white_king;adjacent_cell;cell-has-black_king] [k=4]
  (BKhA) Bool[And(cell-attaked-by-black_king,Not(cell-attaked-by-white))] [k=4]
  (A) Num[And(cell-attaked-by-black_king,Not(cell-attaked-by-white_rook))] [k=4]
  (WQadjBK_row) Bool[And(row-has-white_rook,Exists(adjacent_row,row-has-black_king)))] [k=6]
  (WQadjBK_col) Bool[And(col-has-white_rook,Exists(adjacent_col,col-has-black_king)))] [k=6]
  (WQadjBK) Bool[And(cell-has-white_rook,Exists(adjacent_cell,cell-has-black_king)))] [k=6]
  (WQadjWK) Bool[And(cell-has-white_rook,Exists(adjacent_cell,cell-has-white_king)))] [k=6]
\end{Verbatim}

\subsubsection{All Transition Features}
This features reason about how the value of the features can change in all possible successors.
\begin{Verbatim}[fontsize=\footnotesize]
  (CA) Bool[can(A, DEC)] [k=2]
  (CCM) bool[can(CM, INC)] [k=2]
\end{Verbatim}

\subsubsection{Feature Description}
$CM$. Checkmate\\
$SM$. Stalemate\\
$KE$. Black king is closed in an edge of the table by the white queen\\
$QbK$. White queen is somewhere between kings, rows or columns\\
$A$. The number of cells attaked by black king and not attacked by white queen\\
$BKhA$. Exist a cell attaked by Black king and no attacked by white pieces\\
$Qaw$. White queen is attaked and without protection, i.e. $WQadjBK$ AND $WQadjWK$\\
$WQadjBK_line$. White queen is in an adjacent line to the black king, i.e. $WQadjBK_row$ OR $WQadjWK_col$\\
$CA$. It is possible to decrease the number of cells attaked by black king and not attacked by white queen\\
$CCM$. It is possible to do checkmate in the current state

\subsubsection{Policy}
  \begin{Verbatim}[fontsize=\footnotesize]
   1. KE=0, CA>0 -> {A ↓}
   2. KE=0, CA=0 -> {A ·, KD ·}
   3. BKhA=0, WQadjBK_line>0 -> {KD ·, WQadjBK_line ·}
   4. CCM=0, KE>0 -> {KD ↓}
   5. CCM=1 -> {CM ↑}
  \end{Verbatim}

\subsubsection{Policy Implicits}
The next features has to be in that certain value after the transition for classifing the transition as good, taking into account the policy rules.
  \begin{Verbatim}[fontsize=\footnotesize]
   Qaw=0, Qbk>0, SM=0
  \end{Verbatim}

\subsubsection{Policy description}
Rules 1 and 2 cover the cases in which the king is not enclosed in an edge of the file. The first rule states that it is good to take transitions that decrease the number of squares available to the black king by using the queen. The second rule says that if the first rule is not applicable it is good to take transitions that do not alter the number of squares available to the black king, without moving the white king. The third rule refers to those cases where the black king had moved to its only free square, and classifies as good those transitions in which the queen is moved so that it is not stalemate and remains on the line adjacent to the black king. Rule four covers cases in which the black king is locked on an edge by the queen, and it is good to move the white king closer. finally, rule 5 says that if it is possible to checkmate it is good to do so. It should be noted that thanks to the implicit parts of the policy the king will never be left tied up and unprotected, stalemate will never occur and the white king will always be placed and remain in a line that is between the lines of the kings, whether rows or columns.

\subsubsection{Example of Game}
\begin{Verbatim}[fontsize=\footnotesize]
  K.... # K.... # K.... # ..... # ..... # ..... # ..... # ..... # ..... #
  ..Q.. # ..... # ..... # ..K.. # ..K.. # ..K.. # ..... # ..... # ..... #
  ..... # ..... # ..... # ..... # ..... # ..... # ..K.. # ..K.. # ..K.. #
  ..... # Q.... # Q.... # Q.... # Q.... # Q.... # Q.... # Q.... # ...Q. #
  ....k # ....k # ...k. # ...k. # ..k.. # ...k. # ...k. # ....k # ....k #
\end{Verbatim}

@@

@@

@@

\subsection{Discussoin of Results}

\subsection{Discussoin of Results}

>>>>

Generalized planning is concerned with the computation of general policies that solve multiple instances of a planning domain all at once. It has been recently shown that these policies can be computed in two steps: first, a suitable abstraction in the form of a qualitative numerical planning problem (QNP) is learned from sample plans, then the general policies are obtained from the learned QNP using a planner. In this work, we introduce an alternative approach for computing more expressive general policies which does not require sample plans or a QNP planner. The new formulation is very simple and can be cast in terms that are more standard in machine learning: a large but finite pool of features is defined from the predicates in the planning examples using a general grammar, and a small subset of features is sought for separating “good” from “bad” state transitions, and goals from non-goals. The problems of finding such a “separating surface” while labeling the transitions as “good” or “bad” are
jointly addressed as a single combinatorial optimization problem expressed as a Weighted Max-SAT problem. The advantage of looking for the simplest policy in the given feature space that solves the given examples, possibly non-optimally, is that many domains have no general, compact policies that are optimal. The approach yields general policies for a number of benchmark adversarial domains, even when non-determinism exists in the adverary turn. We use a uniform grammar for all the grid domains capable of capturing the domain representations. These policies can be interpreted qualitatively, explained, and it can be shown mathematically that they will solve all possible instances within a class of problems. This is an advantage when making manual policy changes. We also show an example of how the explainable policies produced by our algorithm can be applied to real robots.

@@

The main advantage of the formulation of generalized planning based on QNPs is that it applies to standard relational domains where the pool of (ground) actions change from instance to instance. On the other hand, while the planning domain is assumed to be given, the QNP abstraction is not, and hence it has to be written by hand or learned. This is the approach of Bonet, Francés, and Geffner (2019) where generalized plans are obtained by learning the QNP abstraction from the domain representation and sample plans, and then solving the abstraction with a QNP planner.

In this work, we build on this thread but introduce an alternative approach for computing general policies that is simpler, yet more powerful. The learning problem is cast as a self-supervised classification problem where (1) a pool of features is automatically generated from a uniform grammar applied to the domain predicates, and (2) a small subset of features is sought for separating “good” from “bad” state transitions, and goals from non-goals. The problems of finding the “separating surface” while labeling the transitions as “good” or “bad” are addressed jointly as a single combinatorial optimization task solved with a Weighted Max-SAT solver. The approach yields general policies for a number of benchmark domains. One of the advantages of having explainable policies is that it gives the engineer the possibility to modify them easily if is necessary. For instance, if there are known exceptions within the domain.

@@

A key question in generalized planning is how to represent general plans or policies when the different instances to be solved have different sets of objects and ground actions. One solution is to work with general features (functions) that have well defined values over any state of any possible domain instance, and think of general policies $\pi$ as mappings from feature valuations into abstract actions that denote changes in the feature values (Bonet and Geffner 2018). In this work, we build on this intuition but avoid the introduction of abstract actions (Bonet and Geffner 2021).

@@

\subsection{Space Invaders}

\subsubsection{Description}
This domain is a simplified version of the original Space Invaders in which the Martians cannot shoot and cannot move below the penultimate row. In this domain the agent can $shoot$, move to the $right$ or to the $left$. The agent during his turn can do two actions, while the Martians can do one. Note that if the Martians and the agent have only one action per turn the agent cannot win from all possible states in the game, this is because of the situation where the agent stands in the Martian's column to shoot him on his next turn, but the next turn (the Martian's turn) the Martian can move away, this leads to an infinite loop if both are playing perfectly. That is, the agent must be able to move to the agent's column and then shoot. To win the agent has to kill all the martians, for this he must shoot each of them. The agent kills an opponent if he shoots while being in the same column, and he can only kill one for each shoot. The martians has three possible actions $left$, $right$, and $down$. All the adversaries move in the same direction. The agent cannot be destroyed in any way by the Martians, they can go down to the row above the agent. The agent cannot collide with the Martians as they cannot descend to the row above it. Note that the policy found is optimal and will solve all possible instances of the game. The policy is also optimal for a version of the game where the Martians destroy the agent when they reach the row above him.

@@

@@

%\subsection{Checkmate in n with Rook}
%\subsubsection{Description}

%\subsubsection{Training instances}
%\begin{Verbatim}[fontsize=\footnotesize]
%  # .  k  .  .  K #
%  # .  .  .  .  . #
%  # .  .  .  .  . #
%  # R  .  .  .  . #
%  # .  .  .  .  . #
%\end{Verbatim}

%\subsubsection{Features}
%\begin{Verbatim}[fontsize=\footnotesize]
%  (CM) Atom[checkmate] [k=1]
%  (SM) Atom[stalemate] [k=1]
%  (QbK) Atom[white_rook_between_kings] [k=1]
%  (KE) Atom[black_king_closed_in_edge_by_white_rook] [k=1]
%  (KD) Dist[cell-has-white_king;adjacent_cell;cell-has-black_king] [k=4]
%  (BKhA) Bool[And(cell-attaked-by-black_king,Not(cell-attaked-by-white))] [k=4]
%  (A) Num[And(cell-attaked-by-black_king,Not(cell-attaked-by-white_rook))] [k=4]
%  (WQadjBK_row) Bool[And(row-has-white_rook,Exists(adjacent_row,row-has-black_king)))] [k=6]
%  (WQadjBK_col) Bool[And(col-has-white_rook,Exists(adjacent_col,col-has-black_king)))] [k=6]
%  (WQadjBK) Bool[And(cell-has-white_rook,Exists(adjacent_cell,cell-has-black_king)))] [k=6]
%  (WQadjWK) Bool[And(cell-has-white_rook,Exists(adjacent_cell,cell-has-white_king)))] [k=6]
%\end{Verbatim}

%\subsubsection{All Transition Features}
%This features reason about how the value of the features can change in all possible successors.
%\begin{Verbatim}[fontsize=\footnotesize]
%  (CA) Bool[can(A, DEC)] [k=2]
%  (CCM) bool[can(CM, INC)] [k=2]
%\end{Verbatim}

%\subsubsection{Feature Description}
%$CM$. Checkmate\\
%$SM$. Stalemate\\
%$KE$. Black king is closed in an edge of the table by the white queen\\
%$QbK$. White queen is somewhere between kings, rows or columns\\
%$A$. The number of cells attaked by black king and not attacked by white queen\\
%$BKhA$. Exist a cell attaked by Black king and no attacked by white pieces\\
%$Qaw$. White queen is attaked and without protection, i.e. $WQadjBK$ AND $WQadjWK$\\
%$WQadjBK_line$. White queen is in an adjacent line to the black king, i.e. $WQadjBK_row$ OR $WQadjWK_col$\\
%$CA$. It is possible to decrease the number of cells attaked by black king and not attacked by white queen\\
%$CCM$. It is possible to do checkmate in the current state

%\subsubsection{Policy}
%  \begin{Verbatim}[fontsize=\footnotesize]
%   1. KE=0, CA>0 -> {A ↓}
%   2. KE=0, CA=0 -> {A ·, KD ·}
%   3. BKhA=0, WQadjBK_line>0 -> {KD ·, WQadjBK_line ·}
%   4. CCM=0, KE>0 -> {KD ↓}
%   5. CCM=1 -> {CM ↑}
%  \end{Verbatim}

%\subsubsection{Policy Implicits}
%The next features has to be in that certain value after the transition for classifing the transition as good, taking into account the policy rules.
%  \begin{Verbatim}[fontsize=\footnotesize]
%   Qaw=0, Qbk>0, SM=0
%  \end{Verbatim}

%\subsubsection{Policy description}
%Rules 1 and 2 cover the cases in which the king is not enclosed in an edge of the file. The first rule states that it is good to take transitions that decrease the number of squares available to the black king by using the queen. The second rule says that if the first rule is not applicable it is good to take transitions that do not alter the number of squares available to the black king, without moving the white king. The third rule refers to those cases where the black king had moved to its only free square, and classifies as good those transitions in which the queen is moved so that it is not stalemate and remains on the line adjacent to the black king. Rule four covers cases in which the black king is locked on an edge by the queen, and it is good to move the white king closer. finally, rule 5 says that if it is possible to checkmate it is good to do so. It should be noted that thanks to the implicit parts of the policy the king will never be left tied up and unprotected, stalemate will never occur and the white king will always be placed and remain in a line that is between the lines of the kings, whether rows or columns.

%\subsubsection{Example of Game}
%\begin{Verbatim}[fontsize=\footnotesize]
%  K.... # K.... # K.... # ..... # ..... # ..... # ..... # ..... # ..... #
%  ..Q.. # ..... # ..... # ..K.. # ..K.. # ..K.. # ..... # ..... # ..... #
%  ..... # ..... # ..... # ..... # ..... # ..... # ..K.. # ..K.. # ..K.. #
%  ..... # Q.... # Q.... # Q.... # Q.... # Q.... # Q.... # Q.... # ...Q. #
%  ....k # ....k # ...k. # ...k. # ..k.. # ...k. # ...k. # ....k # ....k #
%\end{Verbatim}

@@

\section{Dealing with real robots}

First it is necessary to have a simulator of the robot's environment discretised. The policy would be learned with the algorithm described in the paper without involving the robot. This policy is learned from a single instance in the simulator, and can then be run on different and larger instances of the domain. Once the policy is learned with the simulator and the algorithm, it is only necessary to translate the state where the robot is into a state that can be executed by the simulator. Then a transition is selected as good using the already learned policy, and the policy returns the action executed in the simulator to take that transition. The robot controller must be able to handle executing these actions in the real world. This process is repeated until the goal is reached.

So, to be able to execute the policies on real robots, it is only necessary to have a simulator and an algorithm to translate the state of the robot into a state executable by the simulator.

For example, with a turtle robot that has to pick up a certain number of packages and take them to a certain place. We only need a grid simulator, where the objects "package", "agent" and "destination" exist. The simulator must be able to execute the actions to move the robot, pick and drop. The strategy that we would learn with the simulator and our algorithm would be the same as the one learned in the "Delivery" domain, in the results section. The robot controller should be able to cope with executing the actions output by the algorithm, and ideally end up in the state that the simulator expects.

Of course, you could give the initial state translated to the simulator and have the learned policy execute a sequence of actions up to the goal. But this method makes us more vulnerable to the noise that the real world produces.

@@

\subsection{Discussion}
